---
title: "Showing Habitat Change using a Sankey diagram"
---

## Setup

Firstly we load in our libraries and the data. I use the [tidyverse](https://www.tidyverse.org/) a lot for data cleaning and visualisation, `dplyr` (data wrangling), `tidyr` (data tidying), `ggplot2` (data visualisation), `stringr` for working with character strings and `readr` (data reading) all belong to the same family of packages. `ggsankey` is separate but complementary, and will need to be installed from github as opposed to CRAN. 

```{r setup, message = FALSE}
# devtools::install_github("davidsjoberg/ggsankey")
library(ggsankey)
library(dplyr)
library(readr)
library(tidyr)
library(stringr)
library(ggplot2)
data <- readr::read_csv("data/habitat_change.csv")
dplyr::glimpse(data)
```

Whenever I read a dataset into R, I tend to use `dplyr::glimpse` to get a look at the data. There are various ways to do this in R Studio, but I find glimpse to be the quickest and easiest. This example is a fairly typical of habitat data, one before the transition and one after. This could be an extract from the CEH Land Use Change dataset, or a Biodiversity Net Gain proposal. 

In order for `ggsankey` to run properly, we first need to use `tidyr::uncount` to convert the dataset so that 1 row = 1 unit, in this fictional case 1 row = 1 hectare. This function is similar to `pivot_longer` but works better for this situation. We're also going to tidy up the data so that the presentation is easier to digest for non-data audiences. `before` and `after` are useful column names for me, they're description, lower case and 1 word, meaning they're easy to work with in R. However, they could be far more descriptive for a visualiation, so I have changed them to `Initial habitat` and `Current habitat`. When putting a space within a column name in R (which is not recommended until you get to the stage to put it in a graph or table) you must label it with back-ticks - ``. We will also convert the habitat to sentance case so that the outputs in our legend are capitalised. 

```{r data_cleaning, echo = TRUE}
data_processed <- data %>% 
  dplyr::group_by(before, after) %>% 
  tidyr::uncount(area) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(`Initial habitat` = stringr::str_to_sentence(before), 
                `Current habitat` = stringr::str_to_sentence(after))

dplyr::glimpse(data_processed)
```

Then we run the `ggsankey::make_long` function, which formats the data specifically to create the graph. Some R packages will do this if they need the data in a specific way. Then we can create our basic sankey graph, showing the transition from one habitat to another. 

```{r sankey, echo = TRUE}
data_long <- data_processed %>% 
  ggsankey::make_long(`Initial habitat`, `Current habitat`)

sankey_plot <- ggplot2::ggplot(data_long, 
                ggplot2::aes(x = x, next_x = next_x, node = node, 
                    next_node = next_node, fill = factor(node))) +
  ggsankey::geom_sankey() +
  ggsankey::theme_sankey(base_size = 16) 
sankey_plot
```

The great thing about `ggplot2` is that you're able to layer more and more custom options on top of each other, allowing you to programmatically find the right look for your graph and then repeat this across other, similar plots. Using your organisations's brand colour scheme, fonts, etc. is a great way to make reports more aligned and professional. However, it is also essential to make sure that your visualisation is accessible to those with colourblindness.

In the absence of a colourblind friendly brand palette, I would recommend using the [Okabe-Ito palette](https://malcolmbarrett.github.io/ggokabeito/) for discrete datasets, and [Viridis](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html) for continous. I've experimented with a lot of palettes and find these two to be the most accessible and stylish. The land uses in my dummy data come directly from the CEH Land Cover Map, and I have actually created a colourblind friendly palette for these categories downloaded as a QGIS QML file [here](https://www.ceh.ac.uk/sites/default/files/2022-08/LCM%20layer%20files.zip). 

We will also make some more quality of life changes. The x axis label is not neccessary due to our informative naming of the columns earlier and we need to rename the legend title. 

```{r, echo = TRUE}
sankey_plot <- sankey_plot +
  ggokabeito::scale_fill_okabe_ito() +
  ggplot2::labs(x = "", fill = "Habitat") 
sankey_plot
```

There we are! A reasonably professional and fairly novel way to show a habitat transition. Obviously there is far more we can do, but for now I would be quite happy to include that within a project report or dashboard. 